Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-20 paper.tex
Revises intro, leaves space for new results

We will use the dynamic of the learning environment itself. In programming courses, students solve programming problems by implementing
their solution in a particular language; in order to do this, they
must type the code, some times compile it, and finally run it to see how or if
it works. This cycle can be repeated many times, and as they are in this cycle,
studens often change their affective state according to their performence.
Data can, then, be non-intrusively collected while students are
working on their assignments and typing on the keyboard and
interacting with the GUI via mouse or other pointing device.
All the dynamics of actually typing the code and for instance
the syntax errors they received from the compiler or interpreter, the
corrections they tried, the elapsed time between each step, the number of attempts,
compiling errors, exceptions, among others, are collected by the ILE
and can be used. Data mining of all this data could
bring us a better understanding of the learning process, and also a better
model of the student state for both the selection of the appropriate
content to be shown next and to evaluate the
learner\'s competencies. But the problem that remains is to how to recognize
the student's affective state so that it can be used by an adaptive system.

In this work, a method for the recognition of affective states through the
mining of keystroke and mouse dynamics data is proposed.
The hypothesis is that students can vary the dynamic of keystrokes according
to their affective state when programming. A correlation of the
user\'s interactions (keyboard and mouse) with an emotional state, has been
established in preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
exercises.
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based
programming platform was initially proposed by us in
\cite{ijcci17}. In this paper we complete the background and
motivation for this research, and update the state of the art, as well
as justify better the methdology used and the way data has been
collected. The platform we presented in that paper originally was
designed to collect keyboard and mouse data 
from the students\' interaction for data analysis. The proposed method
has three main components: a JavaScript library to capture keyboard
and mouse dynamics from a browser-based editor, a pre-processing step
whose output is a feature vector that is sent to a third component,
the classification algorithm.

An experiment was conducted where students solved a series of programming
assignments using the web based learning environment. Using only the
data from the student\'s keyboard and mouse dynamics six affective
states where recognized for each of the attempts at solving the
assignments. Results obtained from the experiment are promising. The
affective states recognized include the most common in novice
programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and
frustration. For this experiment 
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a na\"ive Bayes. The output of each
classifier determined if the student experienced the affective state during the
assignment.
The proposed method could be used in an ensemble with other sensor
channels to improve the non-invasive recognition of the students’ affective
states when they are working on a programming task.
In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming assignment, they are presented with an ESM survey that asks
what they were feeling during their solving of the assignment.

If a relationship could be modeled, ILEs could adapt the instruction using the
same devices required to input the program to the computer. This could be
important because writing programs in a computer is an important learning
activity when learning programming. In study by Lahtinen, Ala-Mutka \& Järvinen
\cite{lahtinen2005study} is reported that students rated ``working alone on programming coursework'' as a more useful situation than lectures.

The structure of this work is organized as follows: next Section
presents a series of works related to the proposed method in this paper; Section
\ref{sec:method} describes the proposed method for the recognition of affective
states in a web based learning environment for the teaching of programming
languages; next in Section \ref{sec:exp} we explain the experimental evaluation of the
proposed method and results.
Finally, we draw some conclusions about our methodology and
experiments.

>>>>

And in order to do that, we will use the dynamic of the learning environment itself. In programming courses, students solve assignments by implementing
their solution in a particular language; in order to do this, they
must type the code, run it directly or by compiling it to check its results. This cycle can be repeated many times, and as they are in this cycle,
studens often change their affective state according to their performence.
Data can, then, be non-intrusively collected while students are
working on their assignments and typing on the keyboard and
interacting with the GUI via mouse or other pointing device.
All the dynamics of actually typing the code and ILE feedback received
by students, the
corrections they tried, the elapsed time between each step, the number of attempts,
compiling errors, exceptions, among others, are collected by the ILE
and can be used. Data mining of all this data could
bring us a better understanding of the learning process, and also a better
model of the student state for both the selection of the appropriate
content to be shown next and to evaluate the
learner\'s competencies. But the problem that remains is to how to recognize
the student's affective state so that it can be used by an adaptive system.

In this work, a method for the recognition of affective states through the
mining of keystroke and mouse dynamics data is analyzed further after
its initial presentation in \cite{ijcci17}.
Since students can vary the dynamic of keystrokes according
to their affective state when programming, a correlation between the
user\'s interactions (keyboard and mouse) with an emotional state has been
established in preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
exercises.
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based
programming platform was initially proposed by us in
\cite{ijcci17}. In this paper we complete the background and
motivation for this research, and update the state of the art, as well
as justify better the methdology used and the way data has been
collected, extending the results and the methods used. The platform we
presented in that paper originally was 
designed to collect keyboard and mouse data 
from the students\' interaction for data analysis. The proposed method
has three main components: a JavaScript library to capture keyboard
and mouse dynamics from a browser-based editor, a pre-processing step
whose output is a feature vector that is sent to a third component,
the classification algorithm.

An experiment was conducted where students solved a series of programming
assignments using the web based learning environment. Using only the
data from the student\'s keyboard and mouse dynamics six affective
states where recognized for each of the attempts at solving the
assignments. Results obtained from the experiment are promising. The
affective states recognized include the most common in novice
programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and
frustration. For the initial experiment 
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a na\"ive Bayes. The output of each
classifier determined if the student experienced the affective state during the
assignment.
% in this experiment we will also use... - JJ
The proposed method could be used in an ensemble with other sensor
channels to improve the non-invasive recognition of the students’ affective
states when they are working on a programming task.

In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming assignment, they are presented with an ESM survey that asks
what they were feeling during their solving of the assignment.

If a relationship between interaction dynamics and affective state could be modeled, ILEs could adapt the instruction using the
same devices required to input the program to the computer, which
could constitute a breakthrough since writing programs in a computer
is the best way to learn programming by actually doing it. A by Lahtinen, Ala-Mutka \& Järvinen
\cite{lahtinen2005study} reports that students rated ``working alone on programming coursework'' as a more useful situation than lectures.

The rest of the paper is structured as follows: next Section
presents a series of works related to the proposed method in this paper; Section
\ref{sec:method} describes the proposed method for the recognition of affective
states in a web based learning environment for the teaching of programming
languages; next in Section \ref{sec:exp} we explain the experimental evaluation of the
proposed method and results.
Finally, we draw some conclusions about our methodology and
experiments.
